### 2021-07-09 docs/manuscript/manuscript.tex
change genetic for search space

\subsection*{Results}
Out of 2061 GRNs selected using the genetic algorithm, we found 714 groups of
network topologies that were able to correctly interpret the morphogen.
Importantly, the main network motif found to generate the target phenotype in
response to the input signal was the type 3 Incoherent Feed Forward Loop (I3-FFL),
which is in agreement with previous theoretical expectations and experimental
observations. Notably, compared to previously reported pattern forming GRN
topologies, we have uncovered a great variety of novel network designs, some of
which might be worth inquiring through synthetic biology methodologies to test
for the ability of network design with minimal regulatory complexity to robustly
interpret a developmental cue.

@@

\subsection*{Genetic algorithm}

The genetic algorithm used to produce the set of gene regulatory networks (GRNs)
consisted in the following steps:

@@

The source code for the genetic algorithm and other parts of analysis is available at
\url{https://github.com/Nesper94/gene-regulatory-networks}.

@@

On the other hand, analysis of the average vertex degree seems to be indicative
that network topologies are likely to withstand numerous topological
modifications (approximately 4) without significant loss in their ability to
generate a striped pattern of gene expression, representing a high fitness
solution to the prescribed task. Additionally, the average path length seems
to suggest that the evolutionary landscape could be easily traversed by
most network topologies without incurring in significant fitness loss as long
as mutational changes results in moves between adjacent neighbors across
the metagraph (Table~\ref{table1}). It is interesting to note that,
although the examined GRNs seem to exhibit, on average, a high robustness to
topological/parameter changes, they generally show a rather high sensitivity to
variation in the size of the morphogenetic field. For instance, we found that
when the size of the morphogenetic field was fixed at 10, 20, 40 and 50 cells
only 4.66\%, 56.82\%, 56.14\% and 41.05\%, respectively, of the examined
GRNs showed a fitness greater than 0.9 (\nameref{S3_Fig}B).

@@

The average complexity index of network topologies was 39.2435 ($\pm 11.6768$)
and was found to be negatively correlated with the abundance ($r_S = -0.316653$,
$\text{p-value} = 2.25161\times10^{-18}$) (Fig~\ref{fig:ab-rob}B).

Hierarchical clustering of spatiotemporal expression profiles showed that GRNs
with topologies that seemed to be very different have indeed very similar
expression dynamics. We expected to find GRNs with the same topology to group
together but instead we found GRNs with different topologies grouped as close
neighbors (Fig~\ref{fig:clustering}A). Intriguingly, some GRNs displaying
very abundant network topologies were grouped together with GRNs displaying
very infrequent network topologies. This seems to indicate that additional
interactions in the networks are likely to represent alternative regulatory
mechanisms for fine tuning gene expression, instead of drastically changing
the expression mechanism, which can be seen in the expression dynamics of
networks with different topologies (\nameref{S4_Fig}).

@@

Clustering of topologies revealed that there are six main groups of topologies
in our set of 2061 GRNs (Fig~\ref{fig:clustering}B).

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig8}
 \caption{\bf Expression dynamics of topology 9.}
 (A) Network topology number 9. Although this topology consists only of two
 genes, it display a striped pattern of gene expression for both of them as
 can be observed in (B). (B) Spatiotemporal expression profile of a GRN
 with topology 9. Gene \textbf{B} is not shown as it was lost in this topology
 in the evolutionary process simulated by the genetic algorithm.
 \label{fig:topol9}
\end{figure}

Although the topology 9 consists of only two genes, its mechanism of expression
is dynamically more complex than that of other topologies such as the simple
I3-FFL; however, and despite the fact that the Gierer-Meinhardt model form
patterns that scale with tissue size~\cite{shoaf1984}, we found that topology 9
was more sensible to changes in the morphogenetic field size than those
topologies based on I3-FFL (\nameref{S2_Table}),
indicating that the additional regulatory interactions need to be fine-tuned
depending on the number of cells in the field.\\

In the mechanism of expression of topology 9 the expression pattern of the
gene A is initially uniform in the
morphogenetic field, but later it is shaped by the morphogen gradient in such a
way that two important events occur, in first place the overall expression
level of the gene A is decreased by the inhibition of the morphogen, releasing
gene C from its repression by gene A; in second place, it appears a
non-homogeneous spatial distribution in which gene A is expressed at higher
levels in the cells at the rightmost part of the morphogenetic field.\\

@@

Robustness is a defining feature of any developmental patterning process,
and our results suggests that the underlying morphogen interpretation
mechanisms (GRNs) are highly accessible through exploration of the
topology space constrained not only by selection for their final steady state,
but also by their ability to buffer noise~\cite{lo_robust_2015, exelby_2021}.

In our study, the most abundant topologies were found forming highly
connected clusters in a so-called neutral network of genotypes,
implying that the corresponding GRN topologies would tend to be robust
to quantitative changes in the strength of regulatory interactions, diffusion and
degradation parameters, as well as to qualitative changes in the topology
itself. In essence, such organization could be seen as an intrinsic evolutionary
property of GRNs that facilitates the coexistence of robustness and
evolvability~\cite{wagner_robustness_2008}. For instance, bridge topologies
in the neutral network connect distinct classes of pattern-forming
GRNs~(\nameref{S6_Fig}), which suggests that an organism could switch
between different GRNs (for example between a Gierer-Meinhardt-like network
to an Incoherent Feed-Forward Loop) with minimal fitness costs. This high-level
property of the neutral network of topologies could thus facilitate the fine-tuning
of the regulatory systems through genetic modifications in order to meet
specific functional requirements under selective pressures.

@@

Although it has been experimentally demonstrated that the Incoherent Feed
Forward Loop type 2 (I2-FFL) is able to generate a striped gene expression
pattern,~\cite{Schaerli2014, Basu2005}, in our study this motif was present in
only 10.92\% of the topologies. This fact may be because genetic algorithms
can not always find global minima but local minima or its surroundings, however,
it is also possible that the I2-FFL motif is not as robust as other motifs and
therefore be less represented.

@@

The cladogram of spatiotemporal expression profiles shows that the genetic
algorithm selected similar expression mechanisms that can be produced by very
different topologies.

>>>>

\subsection*{Results}
Out of 2061 GRNs selected using the search space algorithm, we found 714 groups of
network topologies that were able to correctly interpret the morphogen.
Importantly, the main network motif found to generate the target phenotype in
response to the input signal was the type 3 Incoherent Feed Forward Loop (I3-FFL),
which is in agreement with previous theoretical expectations and experimental
observations. Notably, compared to previously reported pattern forming GRN
topologies, we have uncovered a great variety of novel network designs, some of
which might be worth inquiring through synthetic biology methodologies to test
for the ability of network design with minimal regulatory complexity to robustly
interpret a developmental cue.

@@

\subsection*{Search space algorithm}

The Markov chain Monte Carlo (MCMC)-like algorithm used to produce the set of
gene regulatory networks (GRNs) consisted in the following steps:

@@

The source code for the search space algorithm and other parts of analysis is
available at \url{https://github.com/Nesper94/gene-regulatory-networks}.

@@

On the other hand, analysis of the average vertex degree seems to be indicative
that network topologies are likely to withstand numerous topological
modifications (approximately 4) without significant loss in their ability to
generate a striped pattern of gene expression, representing a high fitness
solution to the prescribed task. The average complexity index of network
topologies was 39.2435 ($\pm 11.6768$) and was found to be negatively
correlated with the abundance ($r_S = -0.316653$, $\text{p-value} =
2.25161\times10^{-18}$) (Fig~\ref{fig:ab-rob}B).
Additionally, the average path length seems
to suggest that the evolutionary landscape could be easily traversed by
most network topologies without incurring in significant fitness loss as long
as mutational changes results in moves between adjacent neighbors across
the metagraph (Table~\ref{table1}). It is interesting to note that,
although the examined GRNs seem to exhibit, on average, a high robustness to
topological/parameter changes, they generally show a rather high sensitivity to
variation in the size of the morphogenetic field. For instance, we found that
when the size of the morphogenetic field was fixed at 10, 20, 40 and 50 cells
only 4.66\%, 56.82\%, 56.14\% and 41.05\%, respectively, of the examined
GRNs showed a fitness greater than 0.9 (\nameref{S3_Fig}B).

@@

Hierarchical clustering of spatiotemporal expression profiles showed that GRNs
with topologies that seemed to be very different have indeed very similar
expression dynamics. We expected to find GRNs with the same topology to group
together but instead we found GRNs with different topologies grouped as close
neighbors (Fig~\ref{fig:clustering}A). Intriguingly, some GRNs displaying
very abundant network topologies were grouped together with GRNs displaying
very infrequent network topologies. This seems to indicate that additional
interactions in the networks are likely to represent alternative regulatory
mechanisms for fine tuning gene expression, instead of drastically changing
the expression mechanism, which can be seen in the expression dynamics of
networks with different topologies (\nameref{S4_Fig}). Additionally, clustering
of topologies revealed that there are six main groups of topologies in our set
of 2061 GRNs (Fig~\ref{fig:clustering}B).

@@

@@

\begin{figure}[!h]
 \includegraphics[width=\textwidth]{figures/results/Fig8}
 \caption{\bf Expression dynamics of topology 9.}
 (A) Network topology number 9. Although this topology consists only of two
 genes, it display a striped pattern of gene expression for both of them as
 can be observed in (B). (B) Spatiotemporal expression profile of a GRN
 with topology 9. Gene \textbf{B} is not shown as it was lost in this topology
 in the evolutionary process simulated by the search space algorithm.
 \label{fig:topol9}
\end{figure}

Despite having only two genes, the regulatory mechanism underlying
the topology 9 generates dynamically more complex expression patterns
compared to other topologies, such as the
I3-FFL; however, and despite the fact that the Gierer-Meinhardt model form
patterns that scale with tissue size~\cite{shoaf1984}, we found that topology 9
was more sensible to changes in the morphogenetic field size than those
topologies based on I3-FFL (\nameref{S2_Table}),
indicating that the additional regulatory interactions need to be fine-tuned
depending on the number of cells in the field.\\

In the mechanism of expression of topology 9 the initial expression
pattern of gene A is uniform across the morphogenetic field, but later
on the morphogen gradient shapes its dynamics in such a way that
two important events occur: firstly, the overall expression level of the
gene A is decreased by the inhibition of the morphogen, releasing
gene C from its repression by gene A; next, gene A tend to display a
non-homogeneous spatial distribution with higher expression levels
being shifted towards the cells at the rightmost part of the morphogenetic
field.\\

@@

Robustness is a defining feature of any developmental patterning process,
and our results suggests that the underlying morphogen interpretation
mechanisms (GRNs) are highly accessible through exploration of the
topology space constrained not only by selection for their final steady state,
but also by their ability to buffer noise~\cite{lo_robust_2015, exelby_2021}.
In our study, the most abundant topologies were found forming highly
connected clusters in a so-called neutral network of genotypes,
implying that the corresponding GRN topologies would tend to be robust
to quantitative changes in the strength of regulatory interactions, diffusion and
degradation parameters, as well as to qualitative changes in the topology
itself. In essence, such organization could be seen as an intrinsic evolutionary
property of GRNs that facilitates the coexistence of robustness and
evolvability~\cite{wagner_robustness_2008}. For instance, bridge topologies
in the neutral network connect distinct classes of pattern-forming
GRNs~(\nameref{S6_Fig}), which suggests that an organism could switch
between different GRNs (for example between a Gierer-Meinhardt-like network
to an Incoherent Feed-Forward Loop) with minimal fitness costs. This high-level
property of the neutral network of topologies could thus facilitate the fine-tuning
of the regulatory systems through genetic modifications in order to meet
specific functional requirements under selective pressures.

@@

Although it has been experimentally demonstrated that the Incoherent Feed
Forward Loop type 2 (I2-FFL) is able to generate a striped gene expression
pattern,~\cite{Schaerli2014, Basu2005}, in our study this motif was present in
only 10.92\% of the topologies. This fact may be because evolutionary algorithms
can not always find global minima but local minima or its surroundings, however,
it is also possible that the I2-FFL motif is not as robust as other motifs and
therefore be less represented.

@@

The cladogram of spatiotemporal expression profiles shows that the search space
algorithm selected similar expression mechanisms that can be produced by very
different topologies.
